Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-16 wpe.tex
some minor edits to wpe section

For the purposes of this paper, one can view entropy as a measure of complexity and unpredictability in a time series.  A high-entropy time series is almost completely unpredictable---and conversely a low entropy signal is often quite predictable.  This can be made more rigorous: Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the Shannon entropy rate is equal to the sum of the positive Lyapunov exponents, $\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby states of the system will diverge with time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more difficult prediction becomes.

Further we can consider the information contained in a single observable, say the \emph{present}, of the system, measured by the entropy. This information can be partitioned in to two components: the information shared with past observations --- the mutual information between the past and present --- and the information in the present which is not contained in the past --- the conditional entropy. The first part is known as \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the \emph{entropy rate}. In theory, the more redundancy in a signal the more predictable it is. The difference between prediction methods is the types of redundancy (linear, non-linear, etc) they can detect and make use of.

>>>>

For the purposes of this paper, one can view entropy as a measure of complexity and unpredictability in a time series.  A high-entropy time series is almost completely unpredictable---and conversely a low entropy signal is often quite predictable.  This can be made more rigorous: Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the Shannon entropy rate is equal to the sum of the positive Lyapunov exponents, $\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby states of the system will diverge with time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the larger the entropy and the more difficult prediction becomes.

Further we can consider the information contained in a single observable, say the \emph{present}, of the system, measured by the entropy. This information can be partitioned in to two components: the information shared with past observations --- the mutual information between the past and present --- and the information in the present which is not contained in the past --- the conditional entropy. The first part is known as \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the aforementioned \emph{entropy rate}. In theory, the more redundancy in a signal the more predictable it is. The difference between prediction methods is the types of redundancy (linear, non-linear, etc) they can detect and make use of.
